---
title: "Investigation of Pregnancy Code Usage in Primary Care in England"
subtitle: "Short Data Report"
authors:
  - name: Rose Higgins
    affiliations:
      name: "University of Oxford"
      department: "Nuffield Department of Primary Care Health Sciences"
      group: "Bennett Institute for Applied Data Science"
    orcid: 0000-0002-5295-4370
    email: rose.higgins@phc.ox.ac.uk
  - name: Milan Wiedemann
    affiliations:
      name: "University of Oxford"
      department: "Nuffield Department of Primary Care Health Sciences"
      group: "Bennett Institute for Applied Data Science"
    orcid: 0000-0003-1991-282X
    email: milan.wiedemann@phc.ox.ac.uk
  - name: Louis Fischer
    affiliations:
      name: "University of Oxford"
      department: "Nuffield Department of Primary Care Health Sciences"
      group: "Bennett Institute for Applied Data Science"
    orcid: 0000-0002-0295-3812
    email: louis.fisher@phc.ox.ac.uk
  - name: Helen Curtis 
    affiliations:
      name: "University of Oxford"
      department: "Nuffield Department of Primary Care Health Sciences"
      group: "Bennett Institute for Applied Data Science"
    orcid: 0000-0003-3429-9576
    email: helen.curtis@phc.ox.ac.uk
bibliography: references.bib
---

```{r load-pkgs}
#| include: false

library(here)
library(fs)
library(tidyverse)
library(patchwork)
library(ggrepel)
library(gt)
```

```{r}
#| label: load-data
#| include: false
#| cache: true

load(here("sdr", "data", "snomed_code_usage.rda"))
load(here("sdr", "data", "snomed_code_dict.rda"))

used_snomed_concept_ids <- unique(snomed_code_usage$snomed_concept_id)

pregnancy_codelists_paths <- dir_ls(
  here("codelists"),
  glob = "*user-VickiPalin-pregnancy*.csv$"
)

pregnancy_codelists <- pregnancy_codelists_paths |>
  map(read_csv, col_types = list(code = "c")) %>%
  bind_rows(.id = "codelist") |>
  mutate(codelist = str_extract(codelist, "(?<=user-VickiPalin-)\\w+"))

pregnancy_codelists_selected <- pregnancy_codelists |>
  filter(
    codelist %in% c(
      "pregnancy_delivery_snomed_reviewed_2",
      "pregnancy_postdel_antenatal_snomed",
      "pregnancy_antenatal_snomed"
      )
    )
```

```{r load-funs}
#| label: load-funs
#| include: false

source(here("sdr", "scripts", "visualise_code_usage.R"))
```

# Introduction

Identifying pregnancy in primary care electronic healthcare records is difficult.
We aim to explore publicaly available data about code usage in primary care in England to understand how pregnancy related events are recorded since 2011.
This report will focus on:

1. changes and variation in code usage since 2011
2. which codes describe most of the pregnancy related clinical events since 2011

# Methods

## Codelists

We converted Read Codes used by @Minassian2019 to SNOMED CT Codes (see <https://github.com/opensafely/uom_pregnancy_tx_pathways>).
In this report we anlysed the code usage of the following three codelists:

- [pregnancy_delivery_snomed_reviewed_2](https://www.opencodelists.org/codelist/user/VickiPalin/pregnancy_delivery_snomed_reviewed_2/39a121e3/) (Version 39a121e3)
- [pregnancy_postdel_antenatal_snomed](https://www.opencodelists.org/codelist/user/VickiPalin/pregnancy_postdel_antenatal_snomed/22287ec9/) (Version 22287ec9)
- [pregnancy_delivery_snomed_reviewed_2](https://www.opencodelists.org/codelist/user/VickiPalin/pregnancy_antenatal_snomed/61667fb8/) (Version 61667fb8)

## Data source

We used publicly available the SNOMED Code Usage in Primary Care data [@NHSDigital2024] to investigate the code usage of pregnancy related codes between 1st August 2011 to 31 July 2023.

# Results

```{r}
#| label: select-codelist-1
#| include: false
#| cache: true

# Select data for one codelist
selected_codelist <- pregnancy_codelists |> 
  filter(codelist == "pregnancy_delivery_snomed_reviewed_2")

# Get list of codes from codelist that were not found in GP records
unused_codes <- selected_codelist |> 
  filter(!code %in% used_snomed_concept_ids) |> 
  pull(code)

# Select data for all codes from codelist that were found in GP records
selected_snomed_code_usage <- snomed_code_usage |>
  filter(snomed_concept_id %in% selected_codelist$code)

# Get list of 5 most used codes from codelist
selected_top_5_codes <- selected_snomed_code_usage |>
  select(snomed_concept_id, usage) |>
  group_by(snomed_concept_id) |>
  mutate(usage_sum = sum(usage)) |>
  ungroup() |>
  select(-usage) |>
  distinct() |>
  slice_max(usage_sum, n = 5) |>
  pull(snomed_concept_id)

```

## Trends

```{r fig-codelist-1}
#| label: fig-codelist-1
#| fig-cap: "Code usage of 'pregnancy_delivery_snomed_reviewed_2' codelist from 1st August 2011 to 31 July 2022. (A) Overall trend of code usage, (B) trends of the 5 most used codes. Each point represents the end date of a yearly interval." 
#| fig-alt: "A line plot showing code usage of 'pregnancy_delivery_snomed_reviewed_2' codelist"

selected_snomed_code_usage <- selected_snomed_code_usage |> 
  group_by(end_date) |> 
  mutate(usage_codelist = sum(usage, na.rm = TRUE)) |> 
  ungroup()

plot_sum_codelist <- selected_snomed_code_usage |> 
  plot_code_usage(
    date_variable = end_date,
    code_usage_count = usage_codelist
    )

plot_top5_codelist <- selected_snomed_code_usage |> 
  filter(snomed_concept_id %in% selected_top_5_codes) |> 
  plot_code_usage(
    date = end_date,
    code_usage = usage,
    colour = snomed_concept_id
    ) + 
  geom_label_repel(
    aes(
      label = ifelse(
        end_date == min(selected_snomed_code_usage$end_date),
        snomed_concept_id,
        ""
        )
      ),
    direction = "y",
    hjust = "right",
    size = 3,
    box.padding = 0.2) +
  guides(color = "none")


plot_codelist <- plot_sum_codelist / plot_top5_codelist

plot_codelist + 
  plot_annotation(tag_levels = c("A")) + 
  plot_layout(axes = "collect") &
  theme(
    plot.tag = element_text(size = rel(1)),
    plot.tag.position = c(-.014, 1.03)
    )
```


## Code counts

```{r}
#| label: tbl-used-codes-codelist-1
#| tbl-cap: Counts of used codes from 'pregnancy_delivery_snomed_reviewed_2' between 1st Aug 2022 and 31 Jul 2023

selected_snomed_code_usage |>
  filter(!snomed_concept_id %in% unused_codes) |>
  filter(end_date == "2023-07-31") |>
  left_join(snomed_code_dict) |>
  select(snomed_concept_id, description, usage) |>
  gt() |>
  cols_label(
    snomed_concept_id = "SNOMED Code",
    description = "Description",
    usage = "Count"
  ) |>
  cols_width(
    snomed_concept_id ~ px(160),
    description ~ px(460),
    usage ~ px(120)
  ) |>
  opt_interactive(
    use_filters = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_resizers= TRUE,
    use_page_size_select = TRUE
    )

```

## Unused codes

```{r}
#| label: tbl-unused-codes-codelist-1
#| tbl-cap: Unused codes from 'pregnancy_delivery_snomed_reviewed_2'

selected_codelist |>
  filter(code %in% unused_codes) |>
  select(-codelist) |>
  gt() |>
  cols_label(
    code = "SNOMED Code",
    term = "Description"
  ) |>
  cols_width(
    code ~ px(160),
    term ~ px(550)
  ) |>
  opt_interactive(
    use_filters = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_resizers= TRUE,
    use_page_size_select = TRUE
    )

```

## Codes with changing descriptions

```{r}
#| include: false

snomed_codes_diff_desc <- selected_snomed_code_usage |>
  filter(!snomed_concept_id %in% unused_codes) |>
  left_join(snomed_code_dict) |> 
  select(snomed_concept_id, description) |> 
  distinct() |> 
  group_by(snomed_concept_id) |> 
  count() |> 
  filter(n > 1) |> 
  pull(snomed_concept_id)
```


There is a total of n = `r length(snomed_codes_diff_desc)` SNOMED codes in the selected codelist with codes that have changing descriptions.

```{r}
#| label: tbl-diff-codes-desc-1
#| tbl-cap: Codes from 'pregnancy_delivery_snomed_reviewed_2' with changing descriptions between 1st Aug 2011 and 31 Jul 2023


selected_snomed_code_usage |>
  left_join(snomed_code_dict) |> 
  filter(snomed_concept_id %in% snomed_codes_diff_desc) |> 
  select(snomed_concept_id, description) |> 
  distinct() |> 
  arrange(snomed_concept_id) |> 
  gt() |>
  cols_label(
    snomed_concept_id = "SNOMED Code",
    description = "Description"
  ) |>
  cols_width(
    snomed_concept_id ~ px(160),
    description ~ px(550)
  ) |>
  opt_interactive(
    use_filters = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_resizers= TRUE,
    use_page_size_select = TRUE
    )
```

## Exploration of active and inactive codes

```{r}
#| label: tbl-codes-active-inactive-1
#| tbl-cap: Number of codes from 'pregnancy_delivery_snomed_reviewed_2' with changing status (active / inactive) between 1st Aug 2011 and 31 Jul 2023

selected_snomed_code_usage |>
  filter(!snomed_concept_id %in% unused_codes) |>
  # left_join(snomed_code_dict) |> 
  select(snomed_concept_id, active_at_start, active_at_end) |> 
  group_by(active_at_start, active_at_end) |> 
  count() |> 
  ungroup() |> 
  mutate(active_at_start = factor(active_at_start, labels = c("No", "Yes")),
         active_at_end = factor(active_at_end, labels = c("No", "Yes"))) |> 
  gt() |>
  cols_label(
    active_at_start = "Active at start",
    active_at_end = "Active at end",
    n = "Count"
  ) |>
  cols_width(
    active_at_start ~ px(250),
    active_at_end ~ px(250),
    n ~ px(100)
  ) |>
  opt_interactive(
    use_filters = FALSE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_resizers= FALSE,
    use_page_size_select = FALSE
    )
```


# Discussion

# References
